Detecting changes in dispersion in COVID-19 incidence time series using a negative binomial model

### 2024-09-25 manuscript/ms.tex
Update ms.tex

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
Understanding patterns in infectious disease incidence is crucial for understanding epidemic dynamics and for developing effective public health responses, but traditional metrics often overlook the inherent variability in how viruses like SARS-CoV-2 (the virus that causes COVID-19) spread. Quantifying this variability (higher-frequency variation around incidence trends) can elucidate important underlying processes such as transmission heterogeneity. 
We developed a method to identify shifts in the degree of dispersion (case clustering) patterns of SARS-CoV-2  within a single time series. We used negative binomial regression to account for factors that might confound the estimation of case count dispersion, allowing us to use a single parameter to infer the degree of dispersion (clustering) of cases. 
We apply the method to incidence data in 144 US counties from the January 1st, 2020 to March 23rd, 2023, and find that conspicuous shifts in dispersion occur across counties concurrently, and that these shifts are not explained by incidence alone. 
Around incidence peaks, dispersion increased spread, with some individuals potentially spreading the virus to many more people than others. 
Additionally, as the pandemic progressed, dispersion also increased, suggesting increased differences in factors such as how people transmit the virus or how susceptible they are to the virus.
Shifts in dispersion can indicate shifts in epidemic phase and information about underlying drivers of transmission, so our method to accurately measure disperison regardless of population size and incidence, provides a way to help public health officials anticipate and manage outbreaks.

@@

By ``metrics of variability'' we mean estimates of the extent to which incidence is dispersed relative to its expected value.
The expectation could be provided by a moving average of past incidence, a process-based model, or another type of population-level model. 
While the expectation itself will vary over time, there may also me important variation in the level of dispersion around the deterministic skeleton of the epidemic. 
In particular, contact tracing data has revealed temporal variation in transmission heterogeneity. 
Working from contacting tracing data, \cite{guo2023statistical} found that the dispersion parameter varied during different phases of the COVID-19 epidemic in Hong Kong, suggesting that variations in the dispersion parameter are important for understanding and managing the population-level dynamics of disease transmission and outbreak control.
Similarly, \cite{ko2023time} found the dispersion parameter varied over time as new variants emerged.

But tradiational contact tracing is very resource intensive, and although new approaches using digital technologies may improve its speed and availability \cite{kretzschmar_impact_2020} there is a need for complementary population-level analyses that can estimate the impacts of heteoregeneity using incidence data, which is more widely available.
Sun et al. \cite{sun_transmission_2021} found a combination of individual-based and population-based strategies was required for SARS-CoV-2 control, highlighting the importance of considering population-level variability and its relationship to individual-level variability. 
Statistical techniques to dispersion in epidemic time series at the population level are emerging.
One area of interest is how variability is related to different phases of an epidemic. For example, the mean and interannual coefficient of variation of measles incidence was used to construct a metric indicative of where a location may be on the path to elimination of a pathogen~\cite{graham_measles_2019}. 
Analyzing variability in terms of bursts of incidence is also important for planning surge capacity in public health systems\cite{wallinga_metropolitan_2018}. 
An index of effective aggregate dispersion (EffDI) was proposed to elucidate clusters of infection directly from incidence data \cite{schneckenreither_assessing_2023}. 
Working from population-level incidence data, \cite{kirkegaard2021superspreading}.
\cite{kirkegaard2021superspreading} estimated the dispersion parameter from population data.

We now discuss in more detail incidence dispersion dynamics, how we measure them, and what they mean.
Infectious disease transmission follows a branching process, as new infections result from exposure to individuals who are currently infectious. Under a wide range of configurations for a branching process model, including the vast majority of parsimoneous ones, the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution (Kendall 1954)

@@

where $\mu_t$ is the expected value for $I_t$ and $\theta_t$ is the dispersion parameter, which is related to the variance by $\mathrm{Var}[I_t] = \mu_t + \mu_t^2 / \theta_t$. Importantly smaller values of the dispersion parameter $\theta_t$ correspond to increasing amounts of dispersion, increasing amounts by which the variance in realized number infected $I_t$ exceeds the expected value, $\mu_t$. Conversely, the distribution of $I_t$ tends to a Poisson distribution (where variance equals mean) as $\theta_t$ becomes large. This model is common in population biology (e.g. May's host-parisitoid work, Grenfell's papers, funk superspreading).

An intepretation of the dispersion parameter for a time series model of counts is that events are $1 + \theta^{-1}$ times as "crowded" in time relative to a Poisson process with the same mean \cite{lloyd_mean_1967} (see appendix). For example $\theta = 1$ corresponds to a situation where the average number of infections in the same timestep as a randomly selected case will exceed the Poisson expectation by a factor of 2. In a simple example relavent to surge capacity in the health care system, $\theta = 1$ implies that a random infectious individual visiting the emergency department at a hospital would find it on average to be twice as crowded with other infectious individuals with the same pathogen than expected for a Poisson process with the same incidence rate.

That covers the interpretation and it is clear that a lot of modeling epidemic dynamics comes down to modeling $\mu_t$ and $\theta_t$. So how do we model $\mu_t$ and $\theta_t$? A time-series approximation to compartment epidemic models uses $\mu_t = \lambda I_{t-1}$ where $\lambda$ is the current local rate at which incidence at time t-1 produces new infections at time $t$ and $\theta_t = I_{t-1}$. Why is $\theta_t = I_{t-1}$? This comes from the assumption that each of the individuals who acquired the infection at $t-1$ form ''independent lineages'' with identically distributed ''local'' rate parameter. Under these conditions $\theta_t = I_{t-1}$ is right for sufficiently large populations. (Kendall 1954, Bjorstand 2002). But this model assumes that supopulations of infectious individuals eminating from one infected individuals are independent and statistically identical to those from another. That is, it assumes susceptible depletion in one lineage does not affect another, transmission rates are equal across lineages, and reporting rates do not vary across lineages.

Since we know that is probably not true, we can use estimates of $\theta_t$ to understand how heterogeneity in transmission, susceptibility and reporting (and other factors) are driving epidemic dynamics. Instead of assuming $\theta_t = I_{t-1}$, let's learn about the relationship and what drives $\theta_t$, first by quantifying how it changes over time. (e.g., does it vary with $I_t$? that's the null hypothesis.) To do so we will also adopt a more general / phenominological model for $\mu_t$. Instead of $\mu_t = \lambda I_{t-1}$, which is a special kind of moving window, we will use a spline which can approximate the solution to a wider range of autoregressive and other models.

Furthermore, the rate of the case-count-generating process is often changing within a timestep, which cannot be accommodated by a Poisson model.

The negative binomial distribution may accurately model a time series if there is a changing process mean within a timestep: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial \cite{cook_notes_nodate}. 
The result is that we should use the dispersion parameter of a negative binomial distribution to measure changes in meaningful case count clustering in settings with differing base population/incidence, not the variance.

\noindent Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion within a timestep that all lead to overdispersion \cite{barron_analysis_1992}.

We developed a method that quantifies the evolution of dispersion along incidence time series, allowing for the detection of changes in clustering that are not due to changes in population size or overall burden of incidence.
We applied the method to COVID-19 incidence data in US counties to investigate the relationships between incidence, dispersion and epidemic dynamic regimes over a portion of the COVID-19 pandemic. 
%From a modern theoretical ecology perspective, investigating beyond the first moment of a process has also been identified as important: ecological experiments are typically geared towards assessing the impacts of the mean strength of causal processes, however the variance about mean effects have been mostly ignored as a driver in biological assemblages, but may be as important as the mean \cite{benedetti-cecchi_importance_2003}.

>>>>

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
Understanding patterns in infectious disease incidence is crucial for understanding epidemic dynamics and for developing effective public health responses. However, traditional metrics used to quantify incidence patterns often overlook variability as an important characteristic in incidence time series. Quantifying higher-frequency variation around incidence trends can elucidate important underlying processes, including transmission heterogeneity or changes in the susceptibility landscape.
We developed a likelihood-ratio testing framework to identify shifts in the degree of dispersion (case clustering) of COVID-19  within a single time series. We used negative binomial regression to account for factors that might confound the estimation of case count dispersion, allowing us to use a single parameter to infer the degree of dispersion of cases. 
We applied the method to COVID-19 incidence data from 144 US counties from January 1st, 2020 to March 23rd, 2023, and found that conspicuous shifts in dispersion occurred across counties concurrently, and that these shifts were not explained by incidence alone. 
Dispersion increased around peaks in incidence such as the major surge in 2022, and dispersion also increased as the pandemic progressed. These increases potentially indicate that some individuals spread SARS-CoV-2 (the virus that causes COVID-19) to many more people than others, that the susceptibility landscape changed, or that there were changes in reporting.
Shifts in dispersion can indicate shifts in epidemic phase as well as provide information about important underlying biological processes. Therefore, our method to estimate dispersion regardless of population size and incidence provides a way to help public health officials anticipate and manage changes in epidemic regime and the drivers of transmission.

@@

By ``metrics of variability'' we mean estimates of the extent to which incidence is dispersed relative to its expected value.
The expectation could be provided by a moving average of past incidence, a process-based model, or another type of population-level model. 
While the expectation itself will vary over time, there may also me important variation in the level of dispersion around the deterministic skeleton of the epidemic. 
Individual-level heterogeneity in transmission scales up to affect population-level dynamics \cite{lloyd-smith_superspreading_2005}, so variability in epidemic trajectories at the population level may provide information about individual-level variability in the transmission process (transmission heterogeneity).
Contact tracing data has revealed temporal variation in transmission heterogeneity, which is quantified using the dispersion parameter of the offspring distribution.
Working from contacting tracing data, \cite{guo2023statistical} found that the dispersion parameter of the offspring distribution varied during different phases of the COVID-19 epidemic in Hong Kong, which is important for understanding and managing the population-level dynamics of disease transmission and outbreak control.
Similarly, \cite{ko2023time} found the dispersion parameter of the offspring distribution varied over time as new variants emerged. 
However, traditional contact tracing is very resource intensive, and although new approaches using digital technologies may improve its speed and availability \cite{kretzschmar_impact_2020} there is a need for complementary population-level analyses that can estimate heteoregeneity using incidence data, which is more widely available.

Sun et al. \cite{sun_transmission_2021} found a combination of individual-based and population-based strategies was required for SARS-CoV-2 control, highlighting the importance of considering population-level variability and its relationship to individual-level variability. Working from population-level incidence data, \cite{kirkegaard2021superspreading}.
\cite{kirkegaard2021superspreading} estimated the offspring distribution dispersion parameter from population data.
Statistical techniques that make inference based on dispersion of epidemic time series at the population level as opposed to dispersion in the offspring distribution are emerging.
One area of interest is how incidence variability is related to different phases of an epidemic. For example, the mean and interannual coefficient of variation of measles incidence was used to construct a metric indicative of where a location may be on the path to elimination of a pathogen~\cite{graham_measles_2019}. 
Analyzing variability in terms of bursts of incidence is also important for planning surge capacity in public health systems\cite{wallinga_metropolitan_2018}. 
An index of effective aggregate dispersion (EffDI) was proposed to elucidate clusters of infection directly from incidence data \cite{schneckenreither_assessing_2023}.

We now discuss in more detail incidence dispersion dynamics (distinct from but related to the dispersion parameter of the offspring distribution), how we measure them, and what they indicate.
Infectious disease transmission follows a branching process, as new infections result from exposure to individuals who are currently infectious. Under a wide range of configurations for a branching process model, including the vast majority of parsimoneous ones, the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution (Kendall 1954).

@@

where $\mu_t$ is the expected value for $I_t$ and $\theta_t$ is the dispersion parameter, which is related to the variance by $\mathrm{Var}[I_t] = \mu_t + \mu_t^2 / \theta_t$. Importantly, smaller values of the dispersion parameter $\theta_t$ correspond to increasing amounts of dispersion, increasing amounts by which the variance in realized number infected $I_t$ exceeds the expected value, $\mu_t$. Conversely, the distribution of $I_t$ tends to a Poisson distribution (where variance equals mean) as $\theta_t$ becomes large. This model is common in population biology (e.g., in the study of measles \cite{grenfell_dynamics_2002}).

An intepretation of the dispersion parameter for a time series model of counts is that events are $1 + \theta^{-1}$ times as "crowded" in time relative to a Poisson process with the same mean \cite{lloyd_mean_1967}. For example, $\theta = 1$ corresponds to a situation where the average number of infections in the same timestep as a randomly selected case will exceed the Poisson expectation by a factor of 2. In a simple example relevent to surge capacity in the healthcare system, $\theta = 1$ implies that a random infectious individual visiting the emergency department at a hospital would find it on average to be twice as crowded with other infectious individuals with the same pathogen than expected for a Poisson process with the same incidence rate.

Therefore, modeling $\mu_t$ and $\theta_t$ is important to the prediction of epidemics. A time-series approximation to compartment epidemic models uses $\mu_t = \lambda I_{t-1}$ where $\lambda$ is the current local rate at which incidence at time $t-1$ produces new infections at time $t$ and $\theta_t = I_{t-1}$. The assertion that $\theta_t = I_{t-1}$ comes from the assumption that each of the individuals who acquired the infection at $t-1$ form ''independent lineages'' with identically distributed ''local'' rate parameter. Under these conditions $\theta_t = I_{t-1}$ is a good approximation for sufficiently large populations (Kendall 1954, Bjorstand 2002). However, this model assumes that supopulations of infectious individuals eminating from one infected individuals are independent and statistically identical to those from another. That is, the model assumes susceptible depletion in one lineage does not affect another, transmission rates are equal across lineages, and reporting rates do not vary across lineages.

In practice, the above assumptions may not hold, so we can use estimates of $\theta_t$ to understand how heterogeneity in transmission, susceptibility and reporting (and other factors) are driving epidemic dynamics. Instead of assuming $\theta_t = I_{t-1}$, drivers of $\theta_t$ can be investigated, first by quantifying how it changes over time and to what degree it is governed by $I_t$. We adopt a more general/phenominological model for $\mu_t$. Instead of $\mu_t = \lambda I_{t-1}$, which is a special kind of moving window, we use a spline which is able to approximate the solution to a wider range of autoregressive and other models.

Furthermore, the rate of the case-count-generating process is often changing within a timestep, which cannot be accommodated by a Poisson model. 
The negative binomial distribution may accurately model a time series if there is a changing process mean within a timestep: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial \cite{cook_notes_nodate}. 
This is an additional reason that the dispersion parameter of a negative binomial distribution should be used to measure changes in meaningful case count clustering in settings with differing base population/incidence, not the variance. Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion within a timestep that all lead to overdispersion \cite{barron_analysis_1992}.

We developed a method that quantifies the evolution of dispersion along incidence time series, allowing for the detection of changes in clustering that are not due to changes in population size or overall burden of incidence.
We applied the method to COVID-19 incidence data in US counties to investigate the relationships between incidence, dispersion and epidemic dynamic regimes over a portion of the COVID-19 pandemic.
